% !TEX root = ../thesis.tex

\chapter{Conclusion}%
\label{cha:conclusion}

This thesis consists of three stand-alone chapters, each of which addresses a
different aspect of my overall aim of using large-scale transaction level data,
insights from behavioural science, and methods from econometrics and machine
learning to study how individuals spend and save money. In this concluding
chapter, I summarise the main findings of each chapter before discussing
the implications for FinTech apps and placing the work into a broader context.

\section{Chapter summaries}

Chapter~\ref{cha:mlbt}, which is joint work with Neil Stewart, introduces a
novel approach to elicit time preferences from mass financial transaction data.
Lab and field studies typically elicit time preferences by making participants
choose between a smaller-sooner and a larger-later effort or payoff. The core
idea of our approach is to identify a series of consumption choices that mimic
this trade-off, and then exploit these choices to estimate time preferences not
only for the subset of users for whom we observe the choice in the data but for
all users in the data.

The three choice scenarios we identify are monthly or yearly payment for car
insurance, purchasing a new mobile phone outright or paying in monthly
installments, and swapping car and home insurer frequently or infrequently. Each of these choices implicitly trades off a higher short-term cost or effort with a lower long-term cost or effort, so that being patient is beneficial in the long-run. We identify individuals in our dataset that
engage in each of these choice scenarios, and then predict their likelihood of
making the patient choice based on features we extract from the data. We call
that likelihood individual's propensity for patience. We then use this
predictive model to predict the propensity for patience also for those users in
our dataset who didn't engage in a particular choice scenario. Finally, we test
whether predicted propensities from difference choice scenarios correlate
positively, as we would expect if they all elicit the same underlying time
preferences.

The main appeal of the approach is that, if successful, it would allow
researchers to estimate time-preferences for a large set of individuals based
on the smaller-sooner and larger-later payoff approach that researchers
commonly use in lab experiments. More broadly, the general approach we
introduce -- identifying scenarios that help us infer a trait, train a predictive model based on users for whom we observe the scenario, and then predict the trait for all users -- could, in principle, be used for other behavioural traits, too.

Large-scale observational data of the type used in this study has
become available only in recent years and is increasingly used across the
social sciences. One challenge with such data is that they do not usually
provide the same detailed information about behavioural traits that can be gained
from observing a small set of people in a lab. Our approach has the potential
to bridge that divide by allowing researchers to augment large-scale
observational data with behavioural traits elicited from the data itself.

We find, however, that propensities for patience based on different choice
scenarios are uncorrelated, suggesting that in our particular application, our
approach is unsuccessful at eliciting underlying preferences.

There are three
ways in which our work could be refined in future research. First, the use of more suitable data: identifying
choice scenarios is made more challenging because of imperfect tagging of
transactions in the data, and because we do not have direct access to the
algorithms used for tagging. Using the approach on data with more complete and
reliable tagging or in collaboration with the data provider, so that anomalies
in the data can be identified and handled, could eliminate that source of risk.
Second, better models: the prediction models we used could be enhanced both with additional
features and by the use of more sophisticated models and further tuning. In
terms of features, we have already discussed that, in our context, adding
variables that more fully capture users' financial constraints might have been
useful. Furthermore, we have relied on a relatively straightforward set of
features, that could be augmented. In terms of models, we have used a
number of relatively simple ``off-the-shelf'' models. Using more advanced
models could help provide more accurate predictions. Finally, it is possible
that the way people actually make intertemporal choices is not well
captured, even approximately, by the discounting models on which our
elicitation approach builds. If this is true, then either eliciting
time-preferences based on a different framework, or focusing on eliciting
different behavioural traits might prove more successful.

Chapter~\ref{cha:entropy}, also join work with Neil Stewart, calculates simple
summary statistics of individuals' spending profiles and tests whether
differences in spending profiles are predictive of the frequency with which
individuals make transactions into their savings accounts. To summarise
spending profiles, we calculate the entropy of an individual's spending profile, which captures the predictability of spending transactions
in a given period.

We focus on spending profiles for three reasons: first, our understanding of
how individuals spend their money is largely based on survey data from a
relatively small number of individuals. Large-scale transaction data of the
type used in this paper has become available to researchers only in recent
years and has not, to the best of our knowledge, been used to explore patterns
in consumer spending behaviour. Second, research suggests that spending
profiles might reflect circumstances in an individual's life that are also
related to their saving behaviour. Spending entropy, defined similarly as in
this chapter, has been found to predict financial distress, calorie intake, and
the frequency of and amount spend during supermarket visits. Finally, we think
of our focus on spending profile as a proof of concept for the broader agenda
of testing whether we can extract simple summary statistics from large-scale
transaction data that capture relevant information about individuals' life
circumstances. If successful, such information could be used in the design of
financial management apps to provide timely and customised assistance to users
when they need it most.

Our results show that spending entropy is predictive of the frequency of
savings transactions, with an effect size similar to that of an increase in
monthly income of between \pounds1,000 and \pounds2,000. This holds true
if we control for simple component parts of entropy, suggesting that the
non-linear way in which entropy combines these components captures something
about spending profiles that is of relevance. The results are also largely
consistent across entropy measures calculated based on different product and
merchant categories. However, the direction of the effect changes for
different types of entropy that differ in the way in which we treat product or
merchant categories in which an individual makes no transactions in a given
period. Exploratory research suggests that the number of such zero count
categories, together with the variation of counts for categories with a
positive number of transactions, goes some way in explaining the outcome. But
more work is necessary to more fully understand this outcome, which we leave
for future research.

Chapter~\ref{cha:eval}, tests whether using Money Dashboard is associated with
a reduction in discretionary spending and an increase in emergency savings. I
use a new estimator proposed by \citet{callaway2021difference} that corrects
for recently identified problems in two-way fixed effects estimates.

I find that users reduce their discretionary spend by between \pounds100 and
\pounds150 (11-17\% of average discretionary spend) once they start using the
app and sustain that reduction throughout the six-month post-signup period I
consider. Looking at disaggregated measures of discretionary spend further
shows that the reduction is the result of maintained month-to-month changes in
behaviour rather than one-off cancellations of direct-debit transactions, that
it results from reducing spending on a number of different categories of
purchases rather than a single one, and that it is a result of changes along
the extensive rather than the intensive margin -- users reduce the number of
transactions they make rather than the value of the average transaction.
Interestingly, users do not seem to use these additional funds to build up
emergency savings: net-inflows into savings accounts do not change after
signup. I can also neither find significant increases in flows into investment
and pension accounts or additional savings accounts that are not linked to the
app, nor additional loan repayments, which suggests that people allocate their
savings to a number of different purposes that we are not powered to detect in
the data.

The main limitation of my approach is that I cannot isolate the effect of MDB
use from possible confounding factors that let users to self-select into using
the app in the first place, and that I cannot isolate the effect of individual
components of the app such as information provision and goal-setting. However, the findings provide novel insight into how people reduce their
spending and are strongly suggestive that using a financial aggregator app can
help people reduce their spending. While -- as my results show -- such a
reduction in spend does not automatically translate into an increase in
emergency savings, they provide an opportunity for such savings. To harness
that opportunity, FinTech apps should experiment with incorporating tools that
make saving easy.

Further research could focus on causally establishing the link between app use
and discretionary spend, as well as testing the effectiveness of various tools
to help users turn those additional funds into savings. Another area that would
be useful to understand is to what extent the effect of app use might depend on
a ``dosage effect'' -- whether the effectiveness of app features is moderated
by the intensity of app use. However, work testing such an effect will have to
isolate dosage of exposure from personal characteristics that make certain
users engage more with the app. Finally, the finding that people may direct
saved funds towards a number of different purposes can be the result of
differences in funds use across people -- different people use funds for
different purposes -- or within people -- people all use the funds for a
variety of different purposes -- or, most plausibly, a combination of the two.
If people spread the additional funds across many different purposes then it
might be the case that they miss out on directing them instead towards the use
with the highest long-term payoff, such as paying downs of credit-card debt or
investing in government-subsidised savings vehicles. This behaviour would be
similar to the finding by \citet{gathergood2019individuals} that -- instead of
first paying down debt of their highest-interest credit card, people tend to
allocate repayments across all their cards in proportion to outstanding debt.
Further research establishing to what extent there is within-person variation
of fund use, and whether such funds are directed towards the most effective
uses would provide further useful insights into savings behaviour, and could
serve as the basis for further development of FinTech apps to help users direct
saved funds towards the most effective users.

\section{Implications for FinTech apps}

My findings also have a number of direct implications for Money Dashboard and
other FinTech apps. First, the finding that users reduce their discretionary
spend significantly when they use Money Dashboard suggests that apps like it do
live up to their promise and help people meet their financial goals and improve
their financial wellbeing. Second, the finding that a reduction in savings does
not automatically translate into additional high-yield savings, together with a
large body of research suggesting that tools like defaults, implementation
intentions, or commitment devices can help people achieve their goals, suggests
that there is untapped potential to make apps more effective. Further research
to rigorously test and refine such interventions both in-house and with
external collaborators could help apps become more effective and thus more
attractive for customers. Finally, and more generally, the findings show that
the collaboration between FinTech apps and researchers is a win-win strategy
for researchers to develop, test, and refine tools based on the latest
theoretical insights at a large-scale and in relevant contexts, and for
apps to benefit from the insights from that research in the form of more
effective and -- ultimately -- more attractive and profitable apps. 

\section{Broader context and recommendations to researchers}

This thesis has sought to contribute to a better understanding of how people make financial decisions and how financial technology and insights from behavioural science can be harnessed to help them make such decisions in ways that are aligned with their personal goals in cases where this is not the case. I believe this is an area or research with a large potential to do good. But I also believe that it is
important to not loose sight of the bigger picture and realise that
behavioural interventions -- however promising -- can only be part of the
solution that must include broader measures. For instance,
\citet{agarwal2017shapes} points out the important role of financial product
design so that customers can understand, evaluate, and compare products, as
well as the important role of financial regulation aimed at consumer
protection. More broadly, \citet{chater2022frame} point out that attempts to
address problems with behavioural solutions are in danger of unwittingly
crowding-out support for more impactful societal solutions. I think financial
wellbeing is a case in point: I have discussed research that shows that people
do not save as much as they themselves would like or think they could, indicating that there is a role for
the kinds of approaches discussed in my thesis to help them do so. However, at
least in some cases, the main problem is not that people do not save
enough of what they earn, but that they do not earn enough to support
themselves and their families in the first place.

It is thus important that research agendas that study the effectiveness of behavioural solutions are balanced by work that studies societal-level policies such as minimum-wage policies, educational
and job-training programs, and improvements in tax schedules. We will not
maximise financial wellbeing if we are content by helping people to better
navigate the world as they find it -- we must also shape that world so that it
is more conducive to financial wellbeing.
