% !TEX root = ../thesis.tex

\section{Theoretical context}%
\label{sec:theoretical_context}
% Explains the theoretical context and rationale for the inquiry underpinning the Thesis.


\subsection{Financial wellbeing and why it matters}%
\label{sub:financial_wellbeing_and_why_it_matters}

Financial wellbeing is a state in which a person can make ends meet in the
present, can feel comfortable about their financial future, can feel
comfortable about money, and has the financial freedom to make choices that
allow for the enjoyment of life. This includes having control over ones
day-to-day and month-to-month finances, not having to borrow to meet
ongoing obligations, being free of debt or being in control of it, having the
capacity to meet unexpected expenses, and -- for a working-age individual --
being on track to build enough savings for retirement.\footnote{The US Consumer
    Financial Protection Bureau defines fiancial wellbeing as ``a state of
    being wherein a person can fully meet current and ongoing financial
    obligations, can feel secure in their financial future, and is able to make
    choices that allow enjoyment of life'' \citep{cfpb2015financial}. The UK's
    Money and Pension service defines current financial wellbeing as ``being
    able to pay the bills and feel comfortable about money'' and longer-term
    financial security as ``having the savings to deal with the expected, the
    unexpected and the longer-term; or having some form of loss protection such
    as home contents or life insurance'' \citep{mps2018building}. Both
    organisations also differentiate some aspects of financial wellbeing
    depending on whether an individual is in their working or retirement age.
Throughout, this discussion, as well as in the analysis in the following
chapters, I focus on working age individuals.}

Financial wellbeing is important because a lack of it can have severe
consequences both in the short-term and in the long-term. In the short-term,
low financial wellbeing means a constant struggle to make ends meet, being
overwhelmed by debt, having circumstances dictate ones of and approach to
money, a higher likelihood of experiencing material hardship -- which
\citep{cfpb2017financial} defines as running out or worrying about running out
of food, not being able to afford medical treatment or a place to live, or have
utilities turned off -- and the need to cut back on essentials for ones
children.\footnote{For more detailed descriptions about the consequences of
    financial hardship, see \citet{cfpb2017financial, mps2018building,
stepchange2017strengthening}.} All of these experiences can cause a decline in
physical and mental health, a loss of productivity, and an overall deteriation
in the quality of life.

In the long-term, the danger is that the situation is not only self-sustaining
but leads to a viscious circle that becomes increasingly hard to escape. This
can happen purely because having to borrow for regular expenses like food and
bills -- as, the Money and Pension service estimates, about 9 million adults in
the UK do \citep{mps2018building} -- can lead to a situation where an
increasing amount has to be spent on servicing dept and thus increses monthly
expenses and the borrowed amounts required to cover them. In addition, however,
this could happen because of increasingly impaired decision making. A
literature on mental scarcity documents that our minds tend to focus on what is
scarce and neglect what is not, concentrating our mental resources where they
are most needed but reducing cognitive bandwith in other domains, which can
lead to poorer decision making.\footnote{See \citet{shah2012some,
    mullainathan2013scarcity, haushofer2014psychology} for excellent summaries
of the litearature.} For instance, \citet{mani2013poverty} find that low-income
shoppers in New Jersey perform worse on cognitive tasks when first promoted to
think about their financial situation while the same prompts had no effect for
wealthier shoppers, and sugarcane farmers in India perform worse on similar
cognitive tasks shortly before the annual harvest (when money is scarce) than
shortly thereafter (when money is plentiful). There is also evidence that
scarcity might lower preductivity in the present: \citet{kaur2021financial}
randomise the timing of wage payments to low-income piece-rate manufacturing
workers and find that workers that receive their wages early and are thus no
longer liquidity constrained make fewer mistakes and increase their output by 7
percent. Hence, if the anxiety caused by having little money to cover current
expenses lowers individuals' productivity in the present and makes it harder to
engage in rational planning for the future -- how to reduce spending, how to
build skills that would help find a higher-paying job, how to transition to a
healthier lifestyle that would help them feel and perform better physically and
mentally -- then low financial wellbeing in the present might beget low
financial wellbeing in the future.

The determinants of financial wellbeing are a combination of circumstantial and
external factors as well as the capabilities, believes, and behaviours of the
individual. \citet{mps2018building} provides a useful categorisation that sees
financial wellbeing as a function of four broad factors: external factors that
include economic conditions, demographics, and other external factors such as a
person's social environment; enablers that include financial confidence and
numeracy, a sense of control, ones spending and savings mindset, and ones
engagement with money, advice, and technology; day-to-day behaviours like
managing the use of credit, avoiding to borrow for everyday spends, active
saving, keeping track of and making adjustments to ones spending, and shopping
around; and, finally, planning ahead behaviours like building financial
resilience through saving and planning for retirement.\footnote{The US Consumer
Financial Protection Bureau uses a similar classification in its definition of
financial wellbeing \citep{cfpb2015financial}.}

While no one factor is deterministic for fiancial wellbeing, research from the
US \citep{cfpb2017financial} and the UK \citet{mps2018building} agrees that, as
expected, external economic factors such as access to education and higher
paying jobs are important. In the UK, for insteance, individuals with an annual
income of \pounds20,000 or below account for 41 percent of the working age
population but for 69 percent of those with less than \pounds100 of savings.
Similarly, 50 percent of people who borrow to cover everyday expenses earn less
than \pounds17,000. So, clearly, a search for and support of effective economic
and social policy measures should be an important part of any effort to improve
societal financial wellbeing. But the same research also shows that higher
incomes are not sufficient. In the UK, 18 percent of individuals with less than
\pounds 100 in savings have a household income of \pounds30,000 or higher, and
20 percent of those who borrow to cover everyday expenses have an income of
\pounds50,000 or higher. In the US, too, there ia large variation in the
characteristics of individuals at each level of financial wellbeing; the
financial wellbeing of the top quarter of people with a high-school degree, for
instance, is higher than that of the bottom half of those with graduate
degrees.

In both countries, research shows that the level of savings is a key
contributor to financial wellbeing. In the US, it is the one factor that
discriminates between different levels of financial wellbeing better than any
other examined factor \citep{cfpb2017financial}. In the UK, it is, together
with behaviour towards credit, the strongest predictor of financial wellbeing
\citep{mps2018building}. In fact, research by a UK charity suggests that having
\pounds1,000 in liquid savings could reduce the probability of being in dept by
almost half. In particular, having a habit of saving regulary -- even more so
than the amounts saved -- has been found to be a key determinant. Other
factors that are posivitely associated with higher financial wellbeing are
confidence in ones ability to achieve ones financial goals, not using debt to
cover everyday expenses, paying ones bills on time, staying withing ones budget
and spending plan, paying credit card balances in full, and checking bank
statements for errors.\footnote{For details on contributors to financial
wellbeing see \citep{cfpb2017financial} for the US and \citet{mps2018building}
for the UK.}

But many people struggle with these behaviours. For instance, in the UK, 21
percent of the working-age population (10.7 million adults) report to raraly or
never save, and 22 percent of the population have less than \pounds100 in
savings, with those holding that amount in a formal savings account being even
lower. Unsurprisingly, then, one in four adults could not pay an unexpected
bill of \pounds300 from their own money. In the US, the situation is similar,
with 30 percent of adults saying they would be unable to cover a bill of \$400.
This is problematic because many households do face unexpected financial shocks
over the course of a year. Also, many households experience high income
volatility: in the US between 2013 and 2018, the median month-to-month change
in household income was 36 percent, with low-income households experiencing
more frequent and larger income dips \citep{jpmorgan2019weathering}. The same
research finds that while families need roughly six weeks of take-home income
in liquid assets to weather a simultaneous income dip and expenditure shock, 65
percent of households lack such a buffer. In line with these findings,
\citet{roll2020income} find that during the coronavirus pandemic, households
with liquid asses of above \$2,000 had significantly lower risk to experience
financial distress (indicators like skipping essential bills, being behind on
credit card debt, being in overdraft) than households with lower savings.

Managing debt is similarly challenging for many. 9 million people in the UK
also borrow to cover expenses for food and bills \citep{mps2018building}, and
many struggle to stay on top of their credit card debt: 2 million cards were in
arrears or default, another 2 million carried persistent debt, and for another
1.6 million cards, owners were persistently making minimum payments only.
Altogether, the study found that 5 million accounts (9 percent of the total)
that were active in January 2015 would, under their current repayment pattern
and without further borrowing, take 10 years to repay their balance
\citep{fca2016credit}. In addition to holding high and persistent balances, a
large body of research also indicates that individuals make other mistakes in
debt management, especially in dealing with credit cards: they choose
suboptimal credit card contracts \citep{agarwal2015consumers}, sometimes
because they are overly susceptible to temporarily low teaser rates
\citep{shui2004time, ausubel1991failure}; they borrow on payday loans before
while being far from having exhausted their credit card limits
\citep{agarwal2009payday}, their repayment amount is overly influenced by
stated minimum payments \citep{sakaguchi2022default}, they pay down debt across
different cards proportionally to outstanding balances instead of prioritising
high-interest cards \citep{gathergood2019individuals}, and they sometimes hold
credit card debt and liquid assets at the same time \citep{gross2002liquidity,
    gathergood2020co}.\footnote{See \citet{agarwal2017shapes} for a more
    complete review of a large body of research documenting consumer choice
inefficiencies and suboptimal financial behaviour.} 


\subsection{Role of behavioural science}%
\label{sub:role_of_behavioural_science}

%%% retirement savings research
- Retirement planning: among working age, 14 pct report to have done
substantial amount of retirement planning, 44 pct at least a good amount. Among
those between 45-65, this is 47 percent. 20 pct of that latter group have no
plan at all. 
- agarwal2017shapes 2.3 for review

Behavioural science can contribute to higher financial wellbeing in three ways.
First, as a discipline it has learned a fair amount about the psychological
factors that drive financial decisions and accumulated a large body of
knowledge about the effectiveness of different solutions that can make it
easier for people to avoid mistakes and achieve their own goals. Second, a
large body of research has used a subset of these insights and solutions to
help people save more for retirement, and it has learned important lessons from
that endeavour, at least some of which translate to other financial behaviours.
Third, behavioural science might be able to help understand and -- ultimately
-- circumvent the barriers that stand in the way of economic and social
policies that shape the financial wellbeing of many people to a large degree
but is largely outside of their control.

Researchers have documented a large number of factors that influence financial
decision making: cognitive limitations and financial literacy
\citep{agarwal2009age, agarwal2013cognitive, korniotis2011older,
    agarwal2010learning, fernandes2014financial, jorring2020financial};
    time-preferences and self-control \citep{frederick2002time,
    read2018intertemporal, ericson2019intertemporal, cohen2020measuring};
    attitude towards money and spending \citep{rick2008tightwads,
    rick2011fatal}; ones perceived locus of control \citep{perry2005control},
    degree of optimism \citep{puri2007optimism}, ability to frame decisions
    broadly rather than narrowly \citep{kumar2008decision}, and propensity to
    gamble \citep{kumar2009gambles}; ones social network
    \citep{bailey2018economic, kuchler2021social}; the degree of financial
    planning \citep{ameriks2003wealth}; and habits
    \citep{blumenstock2018defaults, schaner2018persistent,
        de2013deposit}.\footnote{For two thorough reviews, see
    \citet{agarwal2017shapes} and \citet{greenberg2019financial}.}

Researchers have also developed and tested a large number of approaches aimed
to help people make better decisions. One main area of research here aims to
address limited self-control -- the difficulty most of us face at least
occasionally to act, moment-by-moment, in our own best interest and according
to our own goals. I briefly discuss three such approaches that have the
potential to help people make the fiancial choices they themselves would like
to make.\footnote{For an extensive review of a large number of behavioural
inteventions aimed to address self-control problems, see
\citet{duckworth2018beyond}.} One extensively studied approach are commitment
devices, whereby an individual restricts their future choice set in order to
avoid choosing a self-defeating action. While not everybody makes use of such
devices when offered the opportunity \citep{bryan2010commitment}, and while
they do not work in all contexts \citep{laibson2015don,robinson2018some}, they
have been found to help individuals increase their savings rates
\citep{ashraf2006tying}, quit smoking \citep{gine2010put}, make healthier food
choices \citep{schwartz2014healthier}, and exercise more regularly
\citep{royer2015incentives}.

Another approach are implementation intentions, a particular type of planning
for the achievement of ones goals that involves ``if-then'' intentions, such as
``if I get paid, then I transfer 10 percent of it into my savings account''
\citep{gollwitzer2006implementation, rogers2015beyond}. Such intentions have
been found to support perseverance in pursueing ones goals
\citep{oettingen2010strategies} and to increase overall goal attainment across
different age groups, life domains, and types of obstacles
\citep{gollwitzer2006implementation}.

A third intervention that has the potential to alter financial behaviour is
social norms messaging, whereby people are informed about how their own
behaviour compares with that of a relevant peer group. Such information can be
especially useful for domains where such information is usually not available,
as is the case with spending and saving, and has been successful inducing high
energy use households to lower their energy use without inducing low-use
households to increase theirs \citep{schultz2007constructive,
allcott2011social, allcott2014short, brandon2017effects}. However, the
information can also backfire and has been found to lower participation in
pension savings plans \citep{beshears2015effect} and completion rates of an
online course \citep{rogers2016discouraged}, making it important to test
messages before deploying them on a large scale and to avoid unintended
consequences.

These approaches, together with many others, have the potential to make it
easier for individuals to manage their money and to spend and save in line with
their own goals. To realise that potential, however, we need to be able to test
these approaches on a large scale and at low cost to gather reliable results
and to have the ability to tweak and refine them, and we need to be able to
deply approaches that work on a large scale. The advent of online and mobile
banking and the large amount of data they create enable researchers and
financial institutions to do just that.


\subsection{Large scale data and fintech}%
\label{sub:large_scale_data_and_fintech}


%%% Data allows us to study behav at large scale (which is transforming
%%% soc science research more broadly)

- buyalskaya2021golden for opportunities of new data for social research

- stachl2020predicting show that using sensor and log data from smartphones to
predict peoples' Big Five personality traits is about as accurate as using
social media footprints.

- montjoye2013predicting also predict personality (big five) from mobile log
data and find predictions to be considerably better than random.

- industry uses: Adaptation in financial sector (boe2019machine)


%%% Household finance: used to study household behaviour

- baker2021household


%%% fintech - can help us study user behaviour and implement what works

- Current applications: simpler dashboards (Monzo, Money Dashboard). link to
lit in levi2020mind on improved information display and lowering cost of
information acquisition
- potential for info display: \citet{mckenzie2011misunderstanding} find that people systematically
underestimate the exponential growth, and that showing them difference in long
term outcome from regular savings left to compound makes them more motivated to
save more for retirement. This is something fintech apps can easily help with
by showing projections - or offering to show them - what is my wealth if I save
x, or y, or z.

- robo advising (philippon2019fintech, dacunto2021frontiers)

- Challenges: what is good advise (sabat2019rules)

- \citet{soman2011earmarking} find that earmarking part of salary as savings
increases savings rates. Fintech apps could suggest automatically transferring
fixed amount into labelled savings pot.
- could even be default option - duckworth2018beyond for default lit

- \citet{gargano2021goal} show that setting savings goals in a fintech app (a
soft commitment device \citep{beshears2016beyond}) increases savings.


\citet{guttman2021semblance}
- Aggregator apps an prove particularly useful to study effect of intervention
because we can see complete financial picture, avoiding semblance of success
results. 





%%% privacy concerns
\begin{itemize}
    \item \citet{demontjoye2015unique} find that it's easy to reidentify users
        based on credit card metadata.
\end{itemize}

%%% The subject is understudied

Exception:
\citet{philipps2021supporting}
- Automatic payroll deduction seems to be valued by employees and has helped
people from the squeezed and struggling segments 

This is why I'm focusing on it... Discuss in next section.

Savings understudied. Exceptions: 
beshears2022borrowin


